Graphical representation example: Alternative interpretations

### 2023-11-15 notebooks/Graphical_representation_alternatives.Rmd
Versions section testing the effect of using the "covariance-based" transformation matrix

```{r recompute-oblique-items}
# Oblique case:

## Compute parameters and coordinates:
items_oblique_params <- items_M2PL |> compute_mirt_params(
  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
  cov_matrix = cov_matrix,                              # Input covariances
  dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
)
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = transform_matrix,
  original_coords = FALSE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

@@

>>>>

```{r recompute-oblique-items}
## Compute parameters and coordinates:
items_oblique_params <- items_M2PL |> compute_mirt_params(
  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
  cov_matrix = cov_matrix,                              # Input covariances
  dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
)
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = transform_matrix,
  original_coords = FALSE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

@@

## Using the "covariance-based" transformation matrix

After pondering the results in the previous sections, I have realized the
following: Although the parameters and the coordinates in the original space are
correctly computed, when I transform the coordinates into the rectangular
coordinate system the transformation (i.e. mapping?) matrix I am using is not a
square root of the inverse covariance matrix, as I state in the algorithm for
computing the graphical representation (section 3.4 in the draft). Instead, what
I was using was the square root fo the inverse correlation matrix, therefore
applying a transformation that intends to be "scale invariant". As a result, the
graphical representation shows the parameters scaled by the standard deviation
of each dimension.

However, I foresee that the following will happen: If I use the _correct_ square
root matrix, that is, the square root matrix of the inverse of the covariance
matrix, the graphical representation will be always the same, independently of
the variances of the dimensions (it will only depend on the correlations).

Let's try with the previous two examples, computing first the square root
matrix. Given that $\mathbf{\Sigma} = \mathbf{SRS}$:

$$\mathbf{\Sigma}^{-1} = (\mathbf{SRS})^{-1};$$

$$\mathbf{\Sigma}^{-1} = \mathbf{S}^{-1}\mathbf{R}^{-1}\mathbf{S}^{-1};$$

$$\mathbf{P^TP} = \mathbf{S}^{-1}\mathbf{P^T_RP_R}\mathbf{S}^{-1},$$

where $\mathbf{P_R}$ refers to the (previously used) correlation-based
transformation matrix. Therefore, we can equate[^mat-eq]
$\mathbf{P} = \mathbf{P_R}\mathbf{S}^{-1}$.

[^mat-eq]: They are not necessarily the same, but we can define them to be for
our convenience, as defining it that way allegedly keeps the desired property of
leaving the horizontal axis unrotated.

With this formula, we compute the new (covariance-based) transformation matrix:

```{r recompute-transform-matrix}
cb_transform_matrix <- transform_matrix %*% solve(sd_matrix)
```

Let's verify that this matrix is a square root of the inverse covariance matrix

```{r check-transform-matrix}
t(cb_transform_matrix) %*%
  cb_transform_matrix  |>
  solve()              |>
  all.equal(cov_matrix)
```

Both matrices are the same (with double numerical precision).

Let's now try the latter example with this new transform matrix.

```{r recompute-oblique-coords}
## Compute coordinates:
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = cb_transform_matrix,
  original_coords = FALSE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

```{r item-plot-cov-based-transform-out}
#| fig.cap: |
#|   Item plots in an oblique test space, with variances [0.25, 0.25], and using
#|   the "covariance-based transformation matrix.
plot_oblique + scale_color_manual(values = PALETTE)
```
